\documentclass[10pt]{article}
\usepackage{geometry}
\geometry{letterpaper, margin=0.7in}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{tabularx}
\usepackage{amssymb}

\begin{document}
\begin{center}
    \textbf{\Large FORM US-06}\\[0.5em]
    \textbf{\Large SOCIOECONOMIC DRIVER REPORT}\\[0.2em]
    \textit{Medical Aid in Dying --- Non-Terminal Provision}\\[0.5em]
    \fbox{\parbox{0.8\textwidth}{\centering \small \textbf{HIPAA PRIVACY NOTICE}\\ This document contains Protected Health Information (PHI) as defined under 45 CFR \S~160.103. Unauthorized disclosure is subject to federal civil and criminal penalties.}}
\end{center}

\vspace{0.5em}
\noindent\textbf{Statutory Authority:} State MAiD Act \S~6 (Mandatory Reporting)\\[0.5em]
\noindent\textbf{INSTRUCTIONS:} This form must be completed for all MAiD provisions where the applicant is not terminally ill. It must be transmitted to the State Department of Health within 30 days of provision. Failure to report is a violation of the Act.

\vspace{1em}
\noindent\textbf{\large SECTION A: CASE IDENTIFICATION}
\vspace{0.5em}\\
\noindent\begin{tabularx}{\textwidth}{|X|X|X|}
\hline
APPLICANT FILE ID & DATE OF PROVISION (YYYY-MM-DD) & STATE / COUNTY \\
\vspace{1.5em} & \vspace{1.5em} & \vspace{1.5em} \\
\hline
\end{tabularx}

\vspace{0.5em}
\noindent\textbf{\large SECTION B: CLINICAL ELIGIBILITY BASIS}
\begin{itemize}
    \item[$\square$] \textbf{Stage 1: Decisional Capacity} confirmed by at least TWO independent evaluators.
    \begin{itemize}
        \item[] $\square$ Understanding \quad $\square$ Appreciation \quad $\square$ Reasoning \quad $\square$ Consistent Expression
    \end{itemize}
    \item[$\square$] \textbf{Stage 2: Process Comprehension} demonstrated (safeguards understood and accepted).
    \item[$\square$] Assessment period of 90 days was completed or waived via statutory exception.
    \item[$\square$] Independent coercion screening conducted by unaffiliated social worker/advocate.
    \item[$\square$] Offer of independent legal assistance provided to applicant.
\end{itemize}
\vspace{0.5em}
\noindent PRIMARY DIAGNOSIS (ICD-10 CODE PREFERRED): \hrulefill

\vspace{1em}
\noindent\textbf{\large SECTION C: SOCIOECONOMIC DRIVERS (MANDATORY)}\\[0.2em]
Identify any factor cited by the applicant as a primary or contributing reason for the request.

\vspace{0.5em}
\noindent\textbf{1. HOUSING STATUS} (Check all that apply)
\begin{itemize}
    \item[$\square$] Applicant had stable, adequate housing.
    \item[$\square$] Applicant was unhoused or in emergency shelter.
    \item[$\square$] Applicant resided in housing they deemed inadequate for their disability needs.
    \item[$\square$] Applicant was under eviction notice or imminent risk of homelessness.
\end{itemize}

\noindent\textbf{2. INCOME AND FINANCIAL SECURITY}
\begin{itemize}
    \item[$\square$] Applicant reported sufficient income to meet basic needs.
    \item[$\square$] Applicant reported insufficient income for basic needs (food, shelter, medication).
    \item[$\square$] Applicant cited federal/state disability support (SSI, SSDI) as insufficient.
    \item[$\square$] Applicant cited food insecurity or insufficient SNAP/WIC benefits.
\end{itemize}

\noindent\textbf{3. CARE AND SUPPORT AVAILABILITY}
\begin{itemize}
    \item[$\square$] Applicant had access to required care/support services.
    \item[$\square$] Applicant required services that were UNAVAILABLE in their geographic area.
    \item[$\square$] Applicant required services that were DENIED by insurance (Medicaid/Medicare/Private).
    \item[$\square$] Applicant required services that were financially UNAFFORDABLE out-of-pocket.
    \item[$\square$] Applicant cited social isolation or lack of community connection as a driver.
\end{itemize}

\vspace{1em}
\noindent\textbf{\large SECTION D: ALTERNATIVES PRESENTED \& OUTCOME}\\[0.2em]
For every driver identified in Section C, indicate the outcome of resource presentation.

\vspace{0.5em}
\noindent{\small
\begin{tabularx}{\textwidth}{|X|c|c|c|c|}
\hline
\textbf{RESOURCE CATEGORY} & \textbf{OFFERED \&} & \textbf{ACCEPTED} & \textbf{NOT AVAILABLE} & \textbf{NOT ACCESSIBLE} \\
& \textbf{DECLINED} & \textbf{BUT FAILED} & \textbf{(Geographic)} & \textbf{(Financial)} \\
\hline
Housing Assistance & $\square$ & $\square$ & $\square$ & $\square$ \\
\hline
Income Supplement & $\square$ & $\square$ & $\square$ & $\square$ \\
\hline
Home Care / Support & $\square$ & $\square$ & $\square$ & $\square$ \\
\hline
Pain Management & $\square$ & $\square$ & $\square$ & $\square$ \\
\hline
Psychiatric Care & $\square$ & $\square$ & $\square$ & $\square$ \\
\hline
\end{tabularx}}

\vspace{1em}
\noindent\textbf{\large SECTION E: ATTESTATION --- PROVIDER, APPLICANT, AND WITNESS}\\[0.5em]
\noindent\textbf{Provider Attestation:} I certify that I have presented all available alternatives to the applicant. Where ``Not Available'' or ``Not Accessible'' is selected above, I certify that no such resource exists or is financially accessible to the applicant in this jurisdiction within a clinically relevant timeframe.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{X X}
SIGNATURE OF PROVIDER\hrulefill & DATE SIGNED\hrulefill \\
\end{tabularx}

\vspace{1em}
\noindent\textbf{Applicant Countersignature:} I confirm that the information recorded in Sections C and D accurately reflects the reasons for my request and the outcome of alternatives presented to me.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{X X}
SIGNATURE OF APPLICANT\hrulefill & DATE SIGNED\hrulefill \\
\end{tabularx}

\vspace{1em}
\noindent\textbf{Independent Witness / Coercion Screener:} I was present during the alternatives presentation recorded in Section D, conducted an independent coercion screening, and attest that the applicant's request is free from undue interpersonal influence.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{X X}
SIGNATURE OF WITNESS\hrulefill & DATE SIGNED\hrulefill \\
PRINTED NAME OF WITNESS\hrulefill & WITNESS ROLE / RELATIONSHIP\hrulefill \\
\end{tabularx}

\vspace{1em}
\hrule
\vspace{0.5em}
\noindent\small\textbf{DISTRIBUTION:} Original: State Department of Health (MAiD Secretariat). Copy 1: State Department of Health and Human Services (HHS). Copy 2: County Medical Examiner.
\end{document}
