\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{letterpaper, margin=1in}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{tabularx}
\usepackage{amssymb}

\begin{document}
\begin{center}
    \textbf{\Large FORM MD-06}\\[0.5em]
    \textbf{\Large SOCIOECONOMIC DRIVER REPORT}\\[0.2em]
    \textit{Medical Aid in Dying --- Non-Terminal Provision}\\[0.2em]
    \textbf{PROTECTED B}
\end{center}

\vspace{0.5em}
\noindent\textbf{Statutory Authority:} Act \S~6 (Mandatory Reporting)\\[0.5em]
\noindent\textbf{INSTRUCTIONS:} This form must be completed for all MAiD provisions where the applicant is not terminally ill (Track 2). It must be transmitted to the Ministry within 30 days of provision. Failure to report is a violation of the Act.

\vspace{1em}
\noindent\textbf{\large SECTION A: CASE IDENTIFICATION}
\vspace{0.5em}\\
\noindent\begin{tabularx}{\textwidth}{|X|X|X|}
\hline
APPLICANT FILE ID & DATE OF PROVISION (YYYY-MM-DD) & PROVINCE / TERRITORY \\
\vspace{1.5em} & \vspace{1.5em} & \vspace{1.5em} \\
\hline
\end{tabularx}

\vspace{1em}
\noindent\textbf{\large SECTION B: CLINICAL ELIGIBILITY BASIS}
\begin{itemize}
    \item[$\square$] Applicant demonstrated decisional capacity pursuant to \S~2 of the Act.
    \item[$\square$] Assessment period of 90 days was completed or waived via exception.
\end{itemize}
\noindent PRIMARY DIAGNOSIS (ICD-11 CODE PREFERRED): \hrulefill

\vspace{1em}
\noindent\textbf{\large SECTION C: SOCIOECONOMIC DRIVERS (MANDATORY)}\\[0.2em]
Identify any factor cited by the applicant as a primary or contributing reason for the request.

\vspace{0.5em}
\noindent\textbf{1. HOUSING STATUS} (Check all that apply)
\begin{itemize}
    \item[$\square$] Applicant had stable, adequate housing.
    \item[$\square$] Applicant was unhoused or in emergency shelter.
    \item[$\square$] Applicant resided in housing they deemed inadequate for their disability needs.
    \item[$\square$] Applicant was under eviction notice or imminent risk of homelessness.
\end{itemize}

\noindent\textbf{2. INCOME AND FINANCIAL SECURITY}
\begin{itemize}
    \item[$\square$] Applicant reported sufficient income to meet basic needs.
    \item[$\square$] Applicant reported insufficient income for basic needs (food, shelter, medication).
    \item[$\square$] Applicant cited disability support payments (e.g., ODSP, AISH) as insufficient.
\end{itemize}

\noindent\textbf{3. CARE AND SUPPORT AVAILABILITY}
\begin{itemize}
    \item[$\square$] Applicant had access to required care/support services.
    \item[$\square$] Applicant required services that were UNAVAILABLE in their jurisdiction.
    \item[$\square$] Applicant required services that were WAITLISTED beyond a clinically acceptable timeframe.
    \item[$\square$] Applicant cited social isolation or lack of community connection as a driver.
\end{itemize}

\vspace{1em}
\noindent\textbf{\large SECTION D: ALTERNATIVES PRESENTED \& OUTCOME}\\[0.2em]
For every driver identified in Section C, indicate the outcome of resource presentation.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{|l|c|c|c|}
\hline
\textbf{RESOURCE CATEGORY} & \textbf{OFFERED \& DECLINED} & \textbf{ACCEPTED BUT FAILED} & \textbf{NOT AVAILABLE} \\
\hline
Housing Assistance & $\square$ & $\square$ & $\square$ \\
\hline
Income Supplement / Disability Aid & $\square$ & $\square$ & $\square$ \\
\hline
Home Care / Personal Support & $\square$ & $\square$ & $\square$ \\
\hline
Palliative / Pain Management & $\square$ & $\square$ & $\square$ \\
\hline
Psychiatric / Mental Health Care & $\square$ & $\square$ & $\square$ \\
\hline
\end{tabularx}

\vspace{1em}
\noindent\textbf{\large SECTION E: ATTESTATION --- PROVIDER, APPLICANT, AND WITNESS}\\[0.5em]
\noindent\textbf{Provider Attestation:} I certify that I have presented all available alternatives to the applicant. Where ``Not Available'' is selected above, I certify that no such resource exists or is accessible to the applicant in this jurisdiction within a clinically relevant timeframe.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{X X}
SIGNATURE OF PROVIDER\hrulefill & DATE SIGNED\hrulefill \\
\end{tabularx}

\vspace{1em}
\noindent\textbf{Applicant Countersignature:} I confirm that the information recorded in Sections C and D accurately reflects the reasons for my request and the outcome of alternatives presented to me.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{X X}
SIGNATURE OF APPLICANT\hrulefill & DATE SIGNED\hrulefill \\
\end{tabularx}

\vspace{1em}
\noindent\textbf{Independent Witness:} I was present during the alternatives presentation recorded in Section D and attest that it was conducted as described above.

\vspace{0.5em}
\noindent\begin{tabularx}{\textwidth}{X X}
SIGNATURE OF WITNESS\hrulefill & DATE SIGNED\hrulefill \\
PRINTED NAME OF WITNESS\hrulefill & WITNESS ROLE / RELATIONSHIP\hrulefill \\
\end{tabularx}

\vspace{2em}
\hrule
\vspace{0.5em}
\noindent\small\textbf{DISTRIBUTION:} Original: Ministry of Health (MAiD Secretariat). Copy 1: Ministry of Social Services. Copy 2: Regional Coroner.
\end{document}
